---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rigvf

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Package version: `r packageVersion("rigvf")`<br/>
Compiled date: `r Sys.Date()`

This proof-of-concept uses the ArangoDB REST API to access the Impact
of Genomic Variation on Function ('IGVF') data portal
(<https://data.igvf.org/>).

## Installation

install the development version from [GitHub](https://github.com/)
with:

``` r
## install.packages("pak")
pak::pak("mtmorgan/rigvf")
```

## Example

There is limited functionality implemented, the most useful of which
is the location of variants associated with a particular gene.

```{r example}
rigvf::gene_variants("ENSG00000106633", threshold = 0.85)
```

The help page `?queries` outlines other available user-facing
functions. See `?arango` for more developer-oriented information.
